Contracts for unions and intersections are hard for Nickel

### 2021-05-19 paper.tex
[WIP]Union & intersection

However, this property does hold for \emph{pure} terms, that are terms without
side-effects. In a pure language, or a language with effect tracking like
Haskell\todo{Cite a different language, or remove citation}, pure terms can be
safely identified for applying optimizations. Other languages rely on
under-approximations.

@@

\subsection*{Laziness}

@@

Unions are useful as a fundamental data block.
Examples: nullable, tree-like data structure, multiple representations

\begin{lstlisting}[title={Tree-like data structure}]
type BinaryTree =
Integer | (BinaryTree, BinaryTree)
\end{lstlisting}

\begin{lstlisting}[title={Plain enumeration}]
protocol : `Http | `Ftp | `Sftp
\end{lstlisting}

\begin{lstlisting}[title={Nullable values}]
optionalValue : String | Null
\end{lstlisting}

\begin{lstlisting}[title={Alternative representations}]
person : {name : String, age : Int}
  | {name : String, dateOfBirth : String}
\end{lstlisting}

%   - Intersections are useful for subtyping or refinement (integers/floats
%       e.g)
\begin{lstlisting}[title={Value subtyping}]
1 : Int & Float
\end{lstlisting}
%   - Intersections are useful to encode overloading
\begin{lstlisting}[title={Overloading}]
incr: (Int -> Int) & (Float -> Float) =
  fun x => x + 1
\end{lstlisting}

%   - Intersections are useful to encode mixins (see Scala3)
\begin{lstlisting}[title={Structural mixins}]
Comparator: {cmp: T -> T -> Bool}
Shower: {show: T -> String}

showSmaller : (dict: Comparator & Shower) -> T -> T -> String
\end{lstlisting}
%   - Intersections appears in some form as unions' dual
\begin{lstlisting}[title={Intersection as the dual of unions}]
f : (Num | String) -> Bool
(*@ $\simeq$ @*) f : (Num -> Bool) (*@ $\cup$ @*) (String -> Bool)
\end{lstlisting}

>>>>

In the example above, the first term prints \nickel{"hi"} two times while the
second term prints it only once. However, this property does hold for
\emph{pure} terms, that are terms without side-effects. In a pure language, or a
language with effect tracking like Haskell\todo{Cite a different language, or
remove citation}, pure terms can be safely identified for applying
optimizations. Other languages rely on under-approximations.

@@

@@

\subsection*{Unions}
Unions are fundamental building blocks of programs. A union holds a value whose
shape is known to be in a fixed set of alternatives: for example, a value that
can be either a boolean or a string. They corresponds logically to conjunctions.
Used in combination with products, they can represent arbitrary tree-like
structures, allowing to model a large class of data
(\ref{fig:union-adt-ex}).

\label{fig:union-adt-ex}
\begin{lstlisting}[title={Tree-like data structure}]
type BinaryTree =
Integer | (BinaryTree, BinaryTree)
\end{lstlisting}

Other use-cases for unions include:
\begin{itemize}
    \item To represent nullable values in JSON (\ref{fig:union-nullable-ex}).
    \item For plain enumerations (\ref{fig:union-enum-ex}).
    \item To support different representations of the same data
        (\ref{fig:union-alt-ex}).
\end{itemize}
All of which are useful in a configuration language.

\label{fig:union-nullable-ex}
\begin{lstlisting}[title={Nullable values}]
optionalValue : String | Null
\end{lstlisting}

\label{fig:union-enum-ex}
\begin{lstlisting}[title={Plain enumeration}]
protocol : `Http | `Ftp | `Sftp
\end{lstlisting}

\label{fig:union-alt-ex}
\begin{lstlisting}[title={Alternative representations}]
person : {name : String, age : Int}
  | {name : String, dateOfBirth : String}
\end{lstlisting}

In statically typed languages, and more specifically in functional programming
languages, sums (also called tagged unions, or discriminated unions) are usually
preferred, as constructors for algebraic datatypes. \unsure{This is a very
    misleading statement at best. Sums are "tagged union" sure, but that refers
to an encoding a sums with unions, certainly not an encoding of unions with
sums!} [references]. Sums are unions' cousin where the alternatives are
syntactically separated by an associated discriminating label. Well-known
programming languages with first-class tagged unions include Haskell, OCaml,
Rust, Scala, PureScript, Elm, Swift and many more.

On the other hand, in dynamically typed languages, it is customary to directly
store different kinds of data in one variable without further ceremony. The
runtime system needs to attach type tags to values anyway, which can usually be
observed using functions like JavaScript's \verb+typeof+. Using such bare
unions\resolved{If you can observe with typeof, then it's pretty much a tagged
union, though the tag is implicit} has already been an idiom in dynamically
typed language for a long time. Being able to type and handle these unions is an
important aspect of the migration to a gradual type system. Indeed, a lot of
gradually typed languages either existed originally as dynamically typed
language or aim at backward compatibility with a dynamically typed base
language. This is why despite the presence of a static type system, gradually
typed languages tend to favor bare unions types which is the typed counterpart
of this practice (as e.g in TypeScript \cite{TypeScriptUnions}).

In the case of Nickel, there is no pre-existing untyped language that we have to
support, so we are freer of our choice. However, Nickel is a configuration
language, whose contract and type system is intended to encode data schemas encountered in the wild. Since these schemas do use
untagged unions, there is a strong incentive to support untagged unions at the
type level, or at least, at the contract level.

\subsection*{Intersections}

Intersections are less prevalent. An intersection is a value that belongs to to
two types at the same time. A simple example is a number literal \nickel{1}
which could be accepted as both an \nickel{Int} and a \nickel{Float} without
having to perform conversions(Figure~\ref{fig:intersection-subytping-ex}). In
this example, the intersection \nickel{Int & Float} plays the role of a subtype
of both \nickel{Int} and \nickel{Float}.

\label{fig:intersection-subytping-ex}
\begin{lstlisting}[title={Value subtyping}]
1 : Int & Float
\end{lstlisting}

Intersections of higher-order types corresponding to overloading
(Figure~\ref{fig:intersection-overloading-ex}).

\label{fig:intersection-overloading-ex}
\begin{lstlisting}[title={Overloading}]
incr: (Int -> Int) & (Float -> Float) =
  fun x => x + 1
\end{lstlisting}

They can also be used to combine constraints (such as traits or interfaces), as
done in Scala 3 (Figure~\ref{fig:intersection-mixins-ex}).

\label{fig:intersection-mixins-ex}
\begin{lstlisting}[title={Structural mixins}]
Comparator: {cmp: T -> T -> Bool}
Shower: {show: T -> String}

showSmaller : (dict: Comparator & Shower) -> T -> T -> String
\end{lstlisting}

Finally, intersection naturally show up as the dual of unions. Thus, in a system
featuring unions, a union in a negative positions is a form of intersection
(Figure~\ref{fig:intersection-dual}).

\label{fig:intersection-dual}
\begin{lstlisting}[title={Intersection as the dual of unions}]
f : (Num | String) -> Bool
(*@ $\simeq$ @*) f : (Num -> Bool) (*@ $\cup$ @*) (String -> Bool)
\end{lstlisting}
